---
title: "commands_cheat_sheet"
author: "Katie Ricketts"
date: "2023-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Library
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(here)
library(gtools)
library(RColorBrewer)
library(paletteer)
library(ggthemes)
```


## Loading Data
```{r}
object <- readr::read_csv(file = "data/filename.csv")
```
```{r}
data_frame_name <- read_csv(here("folder", "data_folder", "file_name.csv"))
```
```{r}
?data_name
#get more info about the data
```


## Working Directory
```{r}
getwd()
```
```{r}
here()
#this will tell you where you are, and where you opened the file from
```


## Scientific Notation
```{r}
options(scipen=999)
#cancels scientific notation for the session
```


## Objects
```{r}
object1 <- 1
object2 <- 2
```
```{r}
sum(object1, object2)
```


## Vectors
```{r}
vector <- c(1, 2, 3, 4)
#c stands for concatenate
my_vector_sequence <- c(1:100)
#creates a vector that has the sequence of numbers from 1 to 100
```
```{r}
mean(vector)
median(vector)
sd(vector)
```
```{r}
vector[3]
vector[vector<4]
#pulls out the value in the vector at that position
```


## Classes of Data
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe" #characters always have quotes
my_logical <- FALSE
my_complex <- 2+4i
```
```{r}
class(my_numeric)
#tells you what class of data it is
```
```{r}
is.integer(my_numeric)
#is my_numeric an integer?

my_integer <- as.integer(my_numeric)
#create a new object specified as an integer
```
```{r}
#the class name "dbl" means that the value is a type of numeric (double precision floating point)
```


## NAs
```{r}
is.na(object)
```
```{r}
anyNA(object)
any_na(object)
#these are interchangable
```
```{r}
na.rm=T
#to take out NAs
#put this piece inside the command you are trying to use
```
```{r}
filter(!is.na(variable1))
#filters out the NA values
```
```{r}
data_frame_name %>% 
  summarize(number_nas = sum(is.na(data_frame_name)))
#counts the number of NAs in the entire data frame
data_frame_name %>% 
  summarize(number_nas = sum(is.na(variable1)))
#counts the number of NAs within vaiable1
data_frame_name %>% 
  summarise_all(~(sum(is.na(.))))
#counts the number of NAs across all of the variables
data_frame_name %>% 
  count(variable1, sort = TRUE)
#provides a count of each possible value under variable1, including NA as a possible value
```
```{r}
data_frame_name_tidy <- data_frame_name %>% 
  na_if("-999")
#replace the value -999 with NA in the entire data frame
data_frame_name %>%  
  mutate(new_column = na_if(varible1, "na_value"))
#creates a new column using the values from variable1 in which "na_value" is replaced with NA
```
```{r}
naniar::miss_var_summary(data_frame_name)
#provides a table of all of the variables, the number of NAs in each variable, and the percent of NAs in each variable
```
```{r}
data_frame_name <- readr::read_csv("data/data_frame_name.csv") %>%
  na_if("-999")
read_csv(file = "data/data_frame_name.csv", na = c("-999"))
#these two are interchangeable
#automatically converts all entries of -999 to NA when loading the data
#only do this if you are already familiar with the data and know what is treated as NA
```


## Data Matrices
```{r}
#lab 2.2 for more details
matrix_name <- matrix(vector, nrow=4, byrow=T)
#use vectors and c() to build a data matrix
```
```{r}
colnames(matrix_name) <- character_string_vector1
row_names(matrix_name) <- character_string_vector2
#naming the columns and rows
```
```{r}
rowSums(matrix_name)
#calculations
```
```{r}
cbind(matrix_name, new_column_vector)
rbind(matric_name, new_row_vector)
#adding colummns and rows
```
```{r}
matrix_name[1,2]
#pull out values: [row, column]
matrix_name[1:4]
#pull out elements 1-4 in column 1
matrix_name[ ,2]
#pull out the entire second column
```


## Data Frames
```{r}
#rows are observations
#columns are variables
#cells are values
```
```{r}
data_frame_name <- data.frame(vector1, vector2, vector3)
#building a data frame
```
```{r}
View(data_frame_name)
#pulls up the data frame in another tab
#or you can click on the data frame in the environment panel
```
```{r}
head(data_frame_name, n=10)
#print the first 10 rows of the data frame
tail(data_frame_name, n=10)
#print the last 10 rows of the data frame
data_frame_name[1, 3]
#same operation as with data matrices
data_frame_name$column_name
#calls upon the entire column within the data frame
```
```{r}
data_frame_name <- rbind(data_frame_name, new_row)
#add a new row
data_frame_name$new_column_name <- c(value1, value2, value2)
#add a new column
```
```{r}
table(data_frame_name$variable)
#produces a count of the number of observations in that variable
```
```{r}
filtered_data <- filter(data_frame_name, variable<=10)
```
#### Renaming
```{r}
data_frame_renamed <- rename(data_frame_name, variable1 = "Variable 1", 
                                              variable2 = "Variable.2", 
                                              variable3 = "VAR 3")
```
```{r}
data_frame_name <- select_all(data_frame_name, tolower)
#make all names lowercase
```
```{r}
select_all(data_frame_name, ~str_replace(., " ", "_"))
#get rid of "." "," and spaces
#the last thing listed in the replace() is what the others are being replaced with
```
#### Summary Functions
```{r}
names(data_frame_name)
#column names
dim(data_frame_name)
#dimensions
str(data_frame_name)
#structure
glimpse(data_frame_name)
summary(data_frame_name)
#overview of the data frame
nrow(data_frame_name)
#number of rows
ncol(data_frame_name)
#number of columns
```
#### Factors
```{r}
data_frame_name$variable <- as.factor(data_frame_name$variable)
#turn the variable class into a factor
class(data_frame_name$variable)
#check the class
levels(data_frame_name$variable)
#see the list of possible levels of the factor
```
#### Saving the Data Frame
```{r}
write.csv(data_frame_name, "data_frame_name.csv", row.names = FALSE)
#saving the data
#write the data frame to a csv file
#the row.names=F keeps the row numbers from printing out
```


## dplyr
#### select
```{r}
select(data_frame_name, variable1, variable2, variable3)
#pull out columns of interest (select has a c)
select(data_frame_name, variable1:variable3)
#pull out a range of columns
select(data_frame_name, -variable4)
#pull out every column except variable 4
select(data_frame_name, contains("something"))
#pull out the columns that contain "something" in their names
#can also be done with:
  # starts_with()
  # ends_with()
  # matches() --> columns that match a regular expression
  # one_of() --> columns names that are from a group of names
select(data_frame_name, matches("a.+er"))
#pulls out every column that contains an "a" and ends with "er"
select(data_frame_name, is.character)
#pulls out data that is a character
select_if(data_frame_name, ~!is.numeric(.))
#pulls out the columns that are NOT numeric
#the ! means the opposite of whatever follows (above, it returns TRUE if the value is not numeric)
#sometimes need the ~! sometimes need just !
# "." tells R too look at all of the values in every cell of the data frame
# "~" tells R to just look at the variables (columns)
```
#### filter
```{r}
filter(data_frame_name, variable1 == "character value")
#pull out rows of interest (filter has an r)
#allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal)
filter(data_frame_name, variable1 %in% c(1, 10))
#all rows with a variable1 value between 1 and 10
filter(data_frame_name, between(variable2, 3, 14))
#all rows with variable2 value between 3 and 14
filter(data_frame_name, near(variable3, 4, tol = 0.5))
#all rows with variable3 value within 0.5 units of 4
filter(data_frame_name, variable1 == "character value" & variable2 > 100)
#only the rows that satisfy both conditions
#you can add as many &s as you want, as long as they don't conflict
filter(data_frame_name, variable1 == "character value" | variable2 > 100)
#this "|" acts as "or"
#all rows that satisfy either condition
filter(data_frame_name, xor(condition1, condition2))
#the rows where only one of the condidions is met, and not both
```
#### count
```{r}
data_frame_name %>% 
  count(variable1, variable2, sort = T) # sort=T will arrange in descending order
#provides a count of how many observations are in a column
#acts like a combination of group_by() and n()
```


## Pipes
```{r}
#feed the output of one command into the input of the next command
#the keystroke shortcut is command+shift+M
```
```{r}
data_frame_name %>% 
  mutate_if(is.character, as.factor)
#turn all characters into factors
data_frame_name %>% 
  select(variable1, variable2) %>% 
  filter(variable1 == "character value") %>% 
  arrange(variable1)
#combines select and filter
#the order of commands matters (ex: you can't filter by variable3 if you have already removed variable3)
#arrange will sort the results in ascending order according to the specified variable
#in order to sort in descending order, use arrange(desc(variable1))
```
```{r}
data_frame_name %>% 
  filter(variable1 == "character value") %>% 
  select_if(all_vars(.=="TRUE"))
#this will only show the columns that have a value of TRUE for this specific observation (row)
```
#### Mutate
```{r}
fish %>% 
  mutate(variable_new = variable1*10) %>% 
  select(variable1, variable_new)
#creates a new column from existing columns
```
```{r}
data_frame_name %>%
  mutate_all(tolower)
#make all entries lowercase
data_frame_name %>% 
  mutate(across(c("variable1", "variable2"), tolower))
#specifies individual columns
```
#### ifelse
```{r}
data_frame_name %>% 
  mutate(variable_new = ifelse(variable1 == -999.00, NA, variable1))
#make a new column, if there is a -999.00 then put NA, and if not -999.00 then transpose the value from the original column
#ifelse() works a little bit like a search and replace
```


## Janitor
```{r}
data_frame_name <- janitor::clean_names(data_frame_name)
```
```{r}
tabyl(data_frame_name, variable1)
#produces a count of each value within the variable, as well as the percentages
#can also be written in a pipe:
data_frame_name %>%
 tabyl(variable1)
```


## skimr
```{r}
skim(variable1)
#provides a data frame with many statistical analyses on that variable
```
```{r}
hist(variable1)
#produces a histogram of the variable
```
#### summarize
```{r}
#or summarise will work too
#produces summary statistics for a given variable in a data frame
```
```{r}
data_frame_name %>% 
  summarize(mean_variable1 = mean(variable1),
            min_variable1 = min(variable1),
            max_variable1 = max(variable1),
            sd_variable1 = sd(variable1),
            total = n())
#produce the mean, min, max, standard deviation, and total number of values of variable1
data_frame_name %>% 
  summarise(n_variable2=n_distinct(variable2))
#produces the number of distinct observation values in variable2
#n_distinct can also be run on its own:
n_distinct(data_frame_name$variable2)
```
```{r}
#other useful summary statistics: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column)
```
```{r}
data_frame_name %>%
  group_by(variable1) %>% #we are grouping by variable1, a categorical variable
  summarize(min_variable2 = min(variable2),
            max_variable2 = max(variable2),
            mean_variable2 = mean(variable2),
            total=n())
#outputs the min, max, mean and total for each group within variable1
```
###### across
```{r}
data_frame_name %>%
  summarize(across(c(variable1, variable2, variable3), n_distinct))
#provides the n_distinct() for all three of the variables listed inside the across(c())
```
```{r}
data_frame_name %>%
  summarize(across(contains("mm"), mean, na.rm=T))
#provides the mean for all variables with "mm" and doesn't include NAs
#helpful for continuous variables
```
```{r}
data_frame_name %>%
  summarise_all(mean, na.rm=T)
#summarize across all variables
```
```{r}
data_frame_name %>%
  summarise(across(!c(variable4, variable5, variable6), 
                   mean, na.rm=T))
#summarize across all variables except variable4, variable5, and variable6
```


## tidyr (Lab 8)
```{r}
#tidy data:
#each variable has its own column
#each observation has its own row
#each value has its own cell
```
#### pivot_longer()
```{r}
data_frame_name %>% 
  pivot_longer(-variable1, #variable1 will not pivot
               names_to = "new_variable", #makes a new column name for the former variables to be moved to
               values_to = "var2" #values (from the old variables being moved) moved to a new column titled var2
               )
data_frame_name_new <- 
  data_frame_name %>% 
  pivot_longer(var8:var76, # a range of columns from var8 to var76
               names_to = "var_new1",
               values_to = "var_new2",
               values_drop_na = TRUE #this will drop the NA's
               )
#the variables being pivoted can be specified in a few ways:
#-c(var1, var2, var3) all variables except var1, var2, and var3 will pivot
#var4:var7 the range of variables from var4 to var7 will pivot
#cols = starts_with("prefix") all variables with the prefix will pivot
data_frame_name %>% 
   pivot_longer(
   cols = starts_with("prefix"),
   names_to = "var_new1",
   names_prefix = "prefix", #this will remove the prefix from the value in the new column
   values_to = "var_new2",
   values_drop_na = TRUE)
```
#### pivot_wider()
```{r}
#when you have an observation scattered across multiple rows
```
```{r}
data_frame_name %>% 
  pivot_wider(names_from = "old_variable", 
              values_from = "value_variable")
#the observations under old_variable will become new columns
data_frame_name %>% 
pivot_wider(names_from = (categorical_var), #new column names come from the categorical_var column
                values_from = c(value_col1, value_col2)) #values come from two separate columns
#this will produce a column for each variable within each category
```
#### separate 
```{r}
data_frame_name %>% 
  pivot_longer(
    var1_v1:var3_v3,
    names_to= c("var_new", "v_new"),
    names_sep="_",
    values_to="variable2"
  )
#separates the variables into two different new columns
```
```{r}
data_frame_name %>% 
  separate(variable1, into= c("var1p1", "var1p2"), sep = "_")
#if values within variable1 contain two values (var1p1_var1p2)
```
```{r}
data_frame_name %>% 
  transform(variable1 = str_split(variable1, ";")) %>%
  unnest(variable1)
#if there is more than one value/observation in a row this will split them into separate rows
```
#### unite
```{r}
data_frame_name %>% 
  unite(var1_var2, "var1", "var2", sep = " ")
#creates a new column that is the combination of two previous columns
```


## ggplot (Lab 9 and 10)
```{r}
#plot= data + geom_ + aesthetics
```
```{r}
ggplot(data=data_frame_name, mapping=aes(x=x_variable, y=y_variable))+geom_boxplot()
#one way to build a ggplot^
ggplot(data_frame_name, mapping=aes(x=x_variable, y=y_variable))+
  geom_point()+
  geom_smooth(method=lm, se=T)
#geom_smooth creates a line of best fit, method=lm means that it will be a line rather than a curve
#the shadow around the regression line is the standard error (se=T)
```
```{r}
data_frame_name %>% 
  ggplot(aes(x=x_variable)) + 
  geom_bar()
#don't specify y because bar plots use the count
#bar plots are good for counts!
data_frame_name %>% 
  filter(variable1=="value") %>%
  ggplot(aes(x=x_variable, y=y_variable))+
  geom_col()
#another way to build a bar plot, but specify both x and y
#you can flip the x and y as well to turn the graph on its side
data_frame_name %>% 
  filter(variable1=="value") %>% 
  ggplot(aes(x=x_variable, y=y_variable))+
  geom_bar(stat="identity")
#a third way to build a bar plot with a specified y variable, but using geom_bar rather than geom_col
#the same plot can be build with geom_col
```
```{r}
#+coord_flip() will flip the x and y variables in the plot
#+scale_y_log10() will scale the y variable logarithmically (can also be done with the x variable and with different scaling)
```
```{r}
#+geom_jitter() does the same thing as +geom_point() (scatterplot) but it accounts for overplotting/overlap
```
#### Aesthetics
```{r}
p <- data_frame_name %>% 
  ggplot(aes(x=x_variable, y=y_variable))
#you can store the data frame as an object and then eperiment with geoms and aesthetics
p + geom_point(aes(x=x_var, y=y_var), size=1.75)+
  labs()+
  theme()
#etc etc
```
```{r}
#bar plots:
geom_bar(position="dodge")
#position="dodge" will put the bars side-by-side instead of stacking them
scale_y_continuous(labels = scales::percent)
#this will scale all the bars to the percent (or proportion)
```
```{r}
+labs(title="Title", 
       x="X-Axis", 
       y="Y-Axis")
```
```{r}
+theme(plot.title=element_text(size=rel(1.25), hjust=0.5))
#rel() changes the relative size of the title
#hjust changes the title position
theme(axis.text.x = element_text(angle = 60, hjust = 1))
#this will tilt the text labelling each bar so that they are easier to read
```
```{r}
ggplot(aes(x=x_variable, fill=group_variable, size=variable2))
#fill will color the data in the plot based on the specified variable 
#use color=group_variable if using a scatter plot/line plot without an area to fill, to do the same thing
#size will adjust the size of points relative to the specified variable
ggplot(aes(x=x_variable, y=y_variable, group=categorical_variables))
#group will make individual plots for each category
```
```{r}
grDevices::colors()
#a list of all the built-in colors available to choose from
```

# Lab 11 stuff:
```{r}
p+theme_linedraw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
#alters the appearance of the legend
```

```{r}
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
#list of the ggthemes available
```
```{r}
p+ theme_igray()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

```{r}
?RColorBrewer
#info about the color brewer
```
```{r}
display.brewer.pal(4,"GnBu")
```
```{r}
#be mindful of color blindness, the website has a filter for color blindness
# http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
```
```{r}
# +`scale_colour_brewer()` is for points  
# +`scale_fill_brewer()` is for fills  
```

```{r}
colors <- paletteer::palettes_d_names
```
```{r}
my_palette <- paletteer_d("ggprism::flames")
```
```{r}
barplot(rep(1,14), axes=FALSE, col=my_palette)
```
```{r}
p+scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point() +
  xlim(0, 4) +
  ylim(1, 6)
#zooms into the specified area on the plot
```

#### Faceting
```{r}
+facet_wrap(.~migratory_strategy, ncol=4)
#creates a ribbon of panels
+facet_grid(migratory_strategy~.)
#creates a grid of panels
# a~b
#a is the rows, b is columns
```
```{r}
ecosphere %>% 
  ggplot(aes(x=log10_mass))+
  geom_density()+
  facet_grid(diet~habitat, scales = "free_y")
```









```{r}
homerange <- homerange %>% 
             mutate(mass_category = case_when(log10_mass <= 1.75 ~ "small",
                                              log10_mass > 1.75 & log10_mass <= 2.75 ~ "medium",
                                              log10_mass > 2.75 ~ "large"))
#creates a new categorical variable based off of values from an existing variable
```


## gtools
```{r}
quartiles <- quantcut(homerange$log10_hra)
table(quartiles)
#produces a table of the quartiles of a specific variable within the data
```







```{r}
range(fisheries_new$year)
```
```{r}
fisheries_new %>% 
  filter(between(year, 2008, 2012))
```



```{r}
homerange %>% count(taxon, sort=T) #sorts by the number of observations
```


Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- 
  USPersonalExpenditure %>% 
  as_tibble() %>% 
  mutate(expenditure = rownames(USPersonalExpenditure))
expenditures
```



8. The column titled `INSTNM` is the institution name. We are only interested in the University of California colleges. Make a new data frame that is restricted to UC institutions. You can remove `Hastings College of Law` and `UC San Francisco` as we are only interested in undergraduate institutions.
```{r}
colleges %>% 
  filter_all(any_vars(str_detect(., pattern = "University of California"))) 
```





