---
title: "Lab 9 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
```

## Seating
1. If you already have a group, it will be helpful to start sitting next to each other.
2. If you don't have a group, hang tight we are working on it.
3. Please setup your computer as normal for today.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

## Warm-up
1. In the data folder there is an epidemiology data set on an outbreak of malaria.
2. `rdt` refers to rapid diagnostic test and they are identified here by age group.
3. Make the data tidy and store them as a new object.
```{r}
malaria <- read.csv("data/malaria.csv") # %>% clean_names()
```
```{r}
malaria
```
```{r}
malaria <- janitor::clean_names(malaria)
```

```{r}
malaria_tidy <- malaria %>% 
  pivot_longer(contains("malaria_"),
               names_to = "malaria_rdt_age_group",
               names_prefix = "malaria_",
               values_to = "malaria_rdt")
malaria_tidy
```
```{r}
# Ledford's solution:
malaria_long <- malaria %>% 
  pivot_longer(cols=starts_with("malaria_rdt"),
               names_to = "age_class",
               values_to = "cases") %>% 
  select(newid, data_date, submitted_date, location_name, province, district, age_class, cases)
malaria_long
```



4. Which district had the highest *total* number of cases on July 30, 2020?
```{r}
malaria_tidy %>% 
  filter(data_date == "2020-07-30" & malaria_rdt_age_group == "tot") %>% 
  group_by(district) %>% 
  summarize(max_cases = max(malaria_rdt, na.rm=T)) %>% 
  arrange(-max_cases)
# This has different (specifically lower) numbers because the "tot" is not the total of all of the cases within the district --> should use the sum() command to total all of the cases (as shown below)
```

```{r}
# Ledford's solution:
malaria_long %>% 
  filter(data_date=="2020-07-30") %>% 
  group_by(district) %>% 
  summarize(tot_cases=sum(cases, na.rm=T)) %>% 
  arrange(-tot_cases)
```



